---
title: "Transformations for the EBP in emdi"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Transformations for the EBP in emdi

The R package emdi allows a range of data transformations for the function ebp 
to get domain specific indicators obtained by Empirical Best Prediction (EBP). 
Since the relies on the normality assumption for the error terms transformations 
may help to achieve the normality. 

With emdi version XX, the following options for the transformation argument in function 
ebp will be available: 
\begin{itemize}
\item \texttt{no}: No transformation 
\item \texttt{log}: Log transformation with a deterministic shift
\item \texttt{box.cox}: Box-Cox transformation with a deterministic shift
\item \texttt{dual}: Dual transformation with a deterministic shift
\item \texttt{log.shift}: Log transformation with an optimized shift
\end{itemize}

While the log transformation does not rely on a transformation parameter, the 
Box-Cox, Dual and Log-shift transformation depend on a transformation parameter 
lambda that can be estimated from the data to find the optimal transformation 
parameter. The estimation approach provided in emdi is the restricted maximum 
likelihood following Gurka (2006). 

A comparison of the various data-driven transformations in the EBP, can be found 
in Rojas et al (2019). 

```{r installation}
install.packages('C:/Users/Ann-Kristin/Documents/emdi_2.0.2.tar.gz', repos = NULL, type="source")
library(emdi)
# Load sample data set
data("eusilcA_smp")
data('eusilcA_pop')
```

## Transformation without transformation parameter

### Log transformation 

The log transformation does not depend on a transformation parameter but the 
vector of the dependent variable is shifted to the positive range by a deterministic 
shift. 

```{r log}
ebp_log <- ebp(fixed = eqIncome ~ gender + eqsize + cash + self_empl +
                   unempl_ben + age_ben + surv_ben + sick_ben + dis_ben + rent + 
                   fam_allow + house_allow + cap_inv + tax_adj, 
               pop_data = eusilcA_pop, pop_domains = "district", 
               smp_data = eusilcA_smp, smp_domains = "district",
               threshold = 10885.33, MSE = FALSE, 
               transformation = 'log')
summary(ebp_log)
```

The transformation is log with a zero shift since there are no negative values in 
the dependent variable in this example. The Shapiro-Wilk test rejects normality 
for both error terms. Additional to the tests shown in the summary, the plot 
method can be used to assess the normality of the error terms.

```{r log_plot}
plot(ebp_log)
```

## Transformation with transformation parameter

### Box-Cox transformation 

The Box-Cox transformation depends on one transformation parameter and is only 
defined for positive y. Therefore, a deterministic shift first shifts the dependent 
variable to the positive range and then the transformation is applied. For the 
estimation of the transformation parameter, an interval for the optimization needs 
to be defined. In emdi, a default option can be chosen which equals an interval 
between -1 and 2. This interval will be reasonable for many applications but it 
can happen that the interval needs to be adjusted. This can be done with a numeric
vector of length two defining the lower and upper limit of the interval, e.g. 
c(-1, 2) for the default interval. 

```{r bc}
ebp_bc <- ebp(fixed = eqIncome ~ gender + eqsize + cash + self_empl +
                   unempl_ben + age_ben + surv_ben + sick_ben + dis_ben + rent + 
                   fam_allow + house_allow + cap_inv + tax_adj, 
               pop_data = eusilcA_pop, pop_domains = "district", 
               smp_data = eusilcA_smp, smp_domains = "district",
               threshold = 10885.33, MSE = FALSE, 
               transformation = 'box.cox', interval = 'default')
summary(ebp_bc)
```


```{r bc_plot}
plot(ebp_bc)
```

### Dual transformation 

The Dual transformation depends on one transformation parameter and is only 
defined for positive y. Therefore, a deterministic shift first shifts the dependent 
variable to the positive range and then the transformation is applied. For the 
estimation of the transformation parameter, an interval for the optimization needs 
to be defined. In emdi, a default option can be chosen which equals an interval 
between 0 and 2 since the Dual transformation does not allow for negative transformation 
parameter. The interval will be reasonable for many applications but it 
can happen that the interval needs to be adjusted. This can be done with a numeric
vector of length two defining the lower and upper limit of the interval, e.g. 
c(0, 2) for the default interval. 

```{r dual}
ebp_dual <- ebp(fixed = eqIncome ~ gender + eqsize + cash + self_empl +
                   unempl_ben + age_ben + surv_ben + sick_ben + dis_ben + rent + 
                   fam_allow + house_allow + cap_inv + tax_adj, 
               pop_data = eusilcA_pop, pop_domains = "district", 
               smp_data = eusilcA_smp, smp_domains = "district",
               threshold = 10885.33, MSE = FALSE, 
               transformation = 'dual', interval = 'default')
summary(ebp_dual)
```


```{r dual_plot}
plot(ebp_dual)
```


### Log-Shift transformation 

The Log-Shift transformation depends on one transformation parameter and is only 
defined for positive y. The transformation parameter is the shift such that there 
is no extra deterministic shift even though the positive scale of y is ensured 
internally.For the estimation of the transformation parameter, an interval for the 
optimization needs to be defined. In emdi, a default option can be chosen where the
interval is based on the total range of y. If $min(y) + 1 <= 1$, the lower limit of the 
interval is $|min(y)| + 1$, otherwise the lower limit is 0. The upper limit is 
$\frac{max(y) - min(y)}{2}$. This interval will be reasonable for many applications but it 
can happen that the interval needs to be adjusted. This can be done with a numeric
vector of length two defining the lower and upper limit of the interval, e.g. 
c(20000, 30000) for the default interval. 

```{r logShift}
ebp_logShift <- ebp(fixed = eqIncome ~ gender + eqsize + cash + self_empl +
                   unempl_ben + age_ben + surv_ben + sick_ben + dis_ben + rent + 
                   fam_allow + house_allow + cap_inv + tax_adj, 
               pop_data = eusilcA_pop, pop_domains = "district", 
               smp_data = eusilcA_smp, smp_domains = "district",
               threshold = 10885.33, MSE = FALSE, 
               transformation = 'log.shift', interval = 'default')
summary(ebp_logShift)
```


```{r logShift_plot}
plot(ebp_logShift)
```
